# When Assumptions Fail, What to Do ? {#sec-chap10}

Every ANOVA and linear mixed model fitted in this book rests on assumptions. Chapter 2 examined those assumptions in depth and showed that their consequences are not equal: independence is critical, homoscedasticity is important, normality is often a minor concern. But what happens when the data genuinely violate one or more of these assumptions in ways that cannot be ignored?

This chapter provides a practical toolkit. It begins with transformations, the classical first response to assumption violations, and examines both their genuine utility and their limitations. It then introduces non-parametric alternatives that make no distributional assumptions at all. It also presents robust ANOVA methods that explicitly downweight the influence of outliers and heavy tails. And it ends with the argument that for many of the situations where assumptions fail most seriously (count data, proportions, survival times), the correct solution is not a patch applied to the normal linear model but a fundamentally different model class: the generalised linear model (GLM).

## Transformations: Log, Square Root, Arcsine and Their Modern Critics

### The Classical Rationale

Transformations have been the standard first response to non-normality and heteroscedasticity in biological data since Fisher's time. The logic is simple: if the response variable $y$ violates the assumptions of ANOVA on its original scale, perhaps a transformed version $f(y)$ will not. The three transformations most commonly encountered in biology are the log, square root, and arcsine transformations.

### The Log Transformation

The log transformation is appropriate when the response is strictly positive and the within-group standard deviation is proportional to the mean, a pattern called **multiplicative variation** or **constant coefficient of variation**. This arises naturally when biological processes operate multiplicatively: cell counts doubling, concentrations following pharmacokinetic decay, body masses scaling allometrically.

$$y' = \log(y) \quad \text{or} \quad y' = \log(y + 1) \text{ if zeros are present}$$

```{r}
#| cache: true
#| fig-cap: "Effect of log transformation on a right-skewed response. Left: raw counts with a long right tail and variance increasing with the mean. Right: log-transformed values with approximately symmetric distribution and homogeneous variance across groups. The log transformation is appropriate when within-group standard deviation is proportional to the group mean."
#| label: fig-log-transform-demo

library(ggplot2)
library(patchwork)

set.seed(42)
n <- 30

# Multiplicative data: sd proportional to mean
group  <- factor(rep(c("Control", "Treatment A", "Treatment B"), each = n),
                 levels = c("Control", "Treatment A", "Treatment B"))
y_raw  <- c(rlnorm(n, meanlog = 2.0, sdlog = 0.8),
            rlnorm(n, meanlog = 2.8, sdlog = 0.8),
            rlnorm(n, meanlog = 3.5, sdlog = 0.8))
y_log  <- log(y_raw)

df_trans <- data.frame(y_raw, y_log, group)

p1 <- ggplot(df_trans, aes(x = group, y = y_raw, fill = group)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 1.5) +
  scale_fill_manual(values = c("#2E86AB", "#F6AE2D", "#E84855")) +
  labs(title = "Raw scale", x = NULL, y = "Response") +
  theme_bw() + theme(legend.position = "none")

p2 <- ggplot(df_trans, aes(x = group, y = y_log, fill = group)) +
  geom_boxplot(alpha = 0.5) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 1.5) +
  scale_fill_manual(values = c("#2E86AB", "#F6AE2D", "#E84855")) +
  labs(title = "Log scale", x = NULL, y = "log(Response)") +
  theme_bw() + theme(legend.position = "none")

p1 + p2
```

The key diagnostic for whether a log transformation is appropriate is whether the within-group standard deviation scales with the within-group mean. A plot of group standard deviation against group mean (Section 5.1) will reveal this pattern. If the relationship is approximately linear through the origin, log-transform. If it is not, the log transformation may not be the right choice.

### The Square Root Transformation

The square root transformation is appropriate for count data where the variance is approximately equal to the mean, the pattern expected from a Poisson distribution:

$$y' = \sqrt{y} \quad \text{or} \quad y' = \sqrt{y + 0.5} \text{ if zeros are present}$$

```{r sqrt-transform}
#| cache: true
set.seed(42)

# Poisson-like count data
y_count <- c(rpois(n, lambda = 3),
             rpois(n, lambda = 8),
             rpois(n, lambda = 20))

df_count <- data.frame(
  y = y_count,
  y_sqrt = sqrt(y_count),
  group = group
)

# Compare residual variance before and after transformation
fit_raw <- aov(y ~ group, data = df_count)
fit_sqrt <- aov(y_sqrt ~ group, data = df_count)

cat("Levene's test on raw counts:\n")
print(car::leveneTest(y ~ group, data = df_count))

cat("\nLevene's test on square root transformed counts:\n")
print(car::leveneTest(y_sqrt ~ group, data = df_count))
```

The Levene's test results demonstrate the variance-stabilising effect of the square root transformation clearly. On the raw count scale, the variances differ significantly across groups ($F_{2,87} = 6.44$, $p = 0.002$) which is a direct consequence of the Poisson-like structure where variance scales with the mean. The control group, with a small mean ($\lambda = 3$), has much less spread than the high group ($\lambda = 20$), producing the heteroscedasticity that would distort a standard ANOVA F test.

After square root transformation, the heteroscedasticity disappears entirely: Levene's test is non-significant ($F_{2,87} = 0.22$, $p = 0.801$) and the variances are now approximately equal across groups. This is the theoretical prediction for Poisson data: the variance-stabilising property of the square root transformation for counts is well established and works reliably when the data are genuinely Poisson or close to it.

Two caveats are worth noting for students. First, the square root transformation only stabilises variance reliably when the counts are not severely overdispersed, that is, when the variance is approximately equal to the mean rather than much larger. If a Levene's test remains significant after square root transformation, overdispersion is likely and a negative binomial GLM (Section 10.4) is the more appropriate solution. Second, as with the log transformation, the square root transformation changes the scale of analysis: differences between group means are now differences in square-root-counts, not differences in raw counts. Back-transforming the estimated means by squaring them gives approximate geometric means on the original count scale, which may or may not be the quantity of biological interest.

### The Arcsine Square Root Transformation

The arcsine square root transformation was historically recommended for proportions and percentages, with the goal of stabilising the variance of binomial data:

$$y' = \arcsin(\sqrt{p})$$

```{r arcsine-transform}
#| cache: true

set.seed(42)

# Simulate binomial proportions across three groups
# with equal true proportions but different sample sizes
# Variance of a binomial proportion is p(1-p)/n
# it depends on p, which is what the arcsine transformation
# is designed to stabilise

n_obs <- 50    # observations per group
n_trials <- 20   # number of trials per observation

# Three groups with different true proportions
true_p  <- c(0.10, 0.45, 0.85)
groups  <- factor(rep(c("Low", "Mid", "High"), each = n_obs),
                  levels = c("Low", "Mid", "High"))

# Simulate observed proportions
props <- c(
  rbinom(n_obs, n_trials, true_p[1]) / n_trials,
  rbinom(n_obs, n_trials, true_p[2]) / n_trials,
  rbinom(n_obs, n_trials, true_p[3]) / n_trials
)

df_asin <- data.frame(
  p      = props,
  p_asin = asin(sqrt(props)),
  group  = groups
)

# Variance of raw proportions per group
cat("Variance of raw proportions by group:\n")
print(tapply(df_asin$p, df_asin$group, var))

# Variance after arcsine transformation
cat("\nVariance of arcsine-transformed proportions by group:\n")
print(tapply(df_asin$p_asin, df_asin$group, var))

# Levene's test on raw proportions
cat("\nLevene's test on raw proportions:\n")
print(car::leveneTest(p ~ group, data = df_asin))

# Levene's test on arcsine-transformed proportions
cat("\nLevene's test on arcsine-transformed proportions:\n")
print(car::leveneTest(p_asin ~ group, data = df_asin))
```

The raw proportions show the classical binomial pattern: the group near 0.10 and the group near 0.85 have low variance (because $p(1−p)/n$ is small when $p$ is near 0 or 1), while the group near 0.45 has the highest variance (because $p(1−p)/n$ is maximised near $p=0.5$). Levene's test on the raw proportions is significant. After the arcsine transformation the three group variances are pulled toward equality and Levene's test becomes non-significant.

### The Modern Critique of Transformations

Despite their long history, transformations have attracted serious criticism, and it is important to understand the limitations __before__ applying them reflexively.

**The arcsine transformation is now largely discredited for proportions.** Warton and Hui [-@Warton2011] showed in an influential paper that the arcsine square root transformation performs poorly compared to directly modelling proportions with a binomial GLM. The transformation was designed for an era before generalised linear models were computationally accessible, and it addresses the wrong problem: rather than transforming the response to meet the assumptions of the normal linear model, the correct approach is to choose a model whose assumptions match the data-generating process. A proportion is generated by a binomial process ? model it with a binomial model.

**Transformations change the question being asked.** When you log-transform a response and analyse it with ANOVA, you are testing hypotheses about group differences on the log scale which correspond to differences in geometric means, not arithmetic means. Back-transforming the estimated differences gives _ratios_, not differences. This is sometimes exactly what you want (a fold-change is biologically natural), but it must be a conscious choice, not an incidental consequence of fixing an assumption violation.

A concrete example makes this concrete. We return to the fertiliser experiment, but now suppose the response is leaf chlorophyll concentration which is a strictly positive measurement with multiplicative variation. We fit ANOVA on both the raw and log-transformed scale and compare what the coefficients actually mean:

```{r transform-question}
#| cache: true
set.seed(42)

n <- 20
group <- factor(rep(c("Control", "Nitrogen", "Phosphorus"), each = n),
                levels = c("Control", "Nitrogen", "Phosphorus"))

# Simulate multiplicative data: nitrogen doubles concentration,
# phosphorus increases it by 50%
conc <- c(
  rlnorm(n, meanlog = log(10), sdlog = 0.3),   # Control:    mean ~10
  rlnorm(n, meanlog = log(20), sdlog = 0.3),   # Nitrogen:   mean ~20
  rlnorm(n, meanlog = log(15), sdlog = 0.3)    # Phosphorus: mean ~15
)

df_conc <- data.frame(conc, log_conc = log(conc), group)

# Model 1: ANOVA on raw scale
fit_raw <- lm(conc ~ group, data = df_conc)
cat("=== Coefficients on raw scale ===\n")
print(coef(fit_raw))

# Model 2: ANOVA on log scale
fit_log <- lm(log_conc ~ group, data = df_conc)
cat("\n=== Coefficients on log scale ===\n")
print(coef(fit_log))

cat("\n=== Back-transformed log coefficients (exp) ===\n")
print(exp(coef(fit_log)))
```

Reading the two outputs side by side reveals the shift in question.

On the **raw scale**, the intercept (11.34) is the estimated mean concentration in the Control group, and the Nitrogen coefficient (8.07) is the estimated *arithmetic difference* in mean concentration between Nitrogen and Control: nitrogen adds approximately 8 units of chlorophyll on average.

On the **log scale**, the intercept is the estimated log-mean of the Control group, and the Nitrogen coefficient (0.55) is the estimated difference in log-means. This is not an arithmetic difference, it is a *log-ratio*. Back-transforming by exponentiating gives $e^{0.55} \approx 1.74$: the Nitrogen group has approximately twice the chlorophyll concentration of the Control group, a **ratio** or **fold-change**.

These are genuinely different questions:

| Scale | Coefficient meaning | Biological statement |
|-------|--------------------|--------------------|
| Raw | $\bar{y}_N - \bar{y}_C \approx 8.07$ | Nitrogen adds 8 units of chlorophyll |
| Log | $\exp(\hat{\beta}_N) \approx 1.74$ | Nitrogen doubles chlorophyll concentration |

__Neither is wrong, but they are not interchangeable__. For chlorophyll concentration, the fold-change interpretation is often more biologically natural: a fertiliser that doubles chlorophyll regardless of baseline concentration is making a multiplicative claim. For a clinical measurement like blood pressure, an arithmetic difference ("reduces blood pressure by 10 mmHg") is usually more directly meaningful than a ratio. _The choice of scale should be driven by which question is scientifically relevant, not by which transformation fixes the residual diagnostics._

**Transformations do not always fix both problems simultaneously.** A transformation chosen to stabilise variance may worsen normality, and vice versa. When both assumptions are violated, there may be no single transformation that addresses both, and a generalised linear model is the cleaner solution.

**The practical recommendation:** use log transformations for multiplicative, positive data where the log scale has a natural biological interpretation (concentrations, fold-changes, ratios). Use square root for counts when overdispersion is modest. Avoid arcsine for proportions, use a binomial GLM instead. And always check whether the transformation has actually fixed the problem by re-examining the residual diagnostics after transforming.


## Non-Parametric Alternatives

Non-parametric tests replace the raw observations with their ranks before analysis. Because ranks are bounded and uniformly distributed, no distributional assumption is needed. The cost is a modest loss of power when the normality assumption actually holds, and a loss of direct interpretability: the tests address questions about rank distributions rather than means.

### Kruskal-Wallis Test

The Kruskal-Wallis test is the non-parametric equivalent of one-way ANOVA. It tests whether the rank distributions of the groups are the same, without assuming normality or homoscedasticity.

The test statistic is:

$$H = \frac{12}{N(N+1)} \sum_{j=1}^{k} \frac{R_j^2}{n_j} - 3(N+1)$$

where $R_j$ is the sum of ranks in group $j$, $n_j$ is the group size, and $N$ is the total number of observations. Under the null hypothesis of identical distributions, $H$ follows approximately a $\chi^2$ distribution with $k - 1$ degrees of freedom.

```{r kruskal-wallis}
#| cache: true
set.seed(42)

# Strongly skewed data, normality assumption clearly violated
y_skew <- c(rlnorm(15, meanlog = 0, sdlog = 1.5),
            rlnorm(15, meanlog = 0.8, sdlog = 1.5),
            rlnorm(15, meanlog = 1.5, sdlog = 1.5))
group <- factor(rep(c("Control", "Low", "High"), each = 15))
df_kw <- data.frame(y = y_skew, group = group)

# Kruskal-Wallis test
kruskal.test(y ~ group, data = df_kw)

# Post-hoc pairwise comparisons after Kruskal-Wallis
# using Dunn's test with Holm correction
# install.packages("dunn.test") if needed
dunn.test::dunn.test(df_kw$y, df_kw$group, method = "holm", kw = FALSE)
```

The Kruskal-Wallis test is the right choice when:

- Sample sizes are small ($n < 10$ per group) and residuals are clearly non-normal.
- The response contains extreme outliers that are genuine biological observations rather than measurement errors.
- The response is an ordinal scale where the numerical values do not have a natural arithmetic interpretation.

It is **not** a universal replacement for ANOVA. I repeat, it is **not** a universal replacement for ANOVA. When the normality assumption is approximately met, ANOVA is more powerful. And when the distributions differ in shape as well as location, one group is right-skewed, another is left-skewed, the Kruskal-Wallis test is testing a mixture of location and shape differences that can be difficult to interpret.

### 10.2.2 Friedman Test

The Friedman test is the non-parametric equivalent of one-way repeated measures ANOVA. It ranks the observations within each block (subject) separately and tests whether the rank distributions differ across conditions.

$$F_r = \frac{12}{nk(k+1)} \sum_{j=1}^{k} R_j^2 - 3n(k+1)$$

where $n$ is the number of blocks (subjects), $k$ is the number of conditions, and $R_j$ is the sum of ranks for condition $j$ across all blocks.

```{r friedman-test}
#| cache: true
#| message: false
#| warning: false
set.seed(42)

n_subj <- 20
k <- 4

# Simulate non-normal within-subject data in long format
df_friedman <- data.frame(
  y = c(rexp(n_subj, rate = 0.50),
            rexp(n_subj, rate = 0.40),
            rexp(n_subj, rate = 0.30),
            rexp(n_subj, rate = 0.25)),
  subject = factor(rep(1:n_subj, times = k)),
  time = factor(rep(1:k, each = n_subj))
)

# Friedman test using long format
friedman.test(y ~ time | subject, data = df_friedman)

# Post-hoc pairwise comparisons after Friedman test
# PMCMRplus expects y, groups, blocks as separate vectors
PMCMRplus::frdAllPairsNemenyiTest(
  y = df_friedman$y,
  groups = df_friedman$time,
  blocks = df_friedman$subject
)
```

The Friedman test should be considered when:

- The within-subject response is clearly non-normal even after transformation.
- The response is ordinal.
- Sample sizes are very small (fewer than 10 subjects).

For larger samples with continuous responses, the repeated measures mixed model from Chapter [-@sec-repeat-mixed] is more powerful and more flexible, and the normality assumption for the residuals is approximately satisfied by the central limit theorem.

### Permutation Tests

Permutation tests are the most general non-parametric approach. Rather than deriving a reference distribution from mathematical theory (the $F$ distribution, the $\chi^2$ distribution), they construct an empirical reference distribution by repeatedly shuffling the data and recomputing the test statistic. The p-value is the proportion of permuted statistics that are as extreme as or more extreme than the observed statistic.

```{r permutation-test}
#| cache: true
library(future)
library(furrr)

plan(multisession)

set.seed(42)

# Observed F statistic
fit_obs <- aov(y ~ group, data = df_kw)
obs_F   <- summary(fit_obs)[[1]][["F value"]][1]

# Permutation distribution: shuffle group labels 4999 times
n_perm <- 4999
perm_F <- future_map_dbl(seq_len(n_perm), function(i) {
  df_perm <- df_kw
  df_perm$group <- sample(df_kw$group)
  summary(aov(y ~ group, data = df_perm))[[1]][["F value"]][1]
}, .options = furrr_options(seed = TRUE))

plan(sequential)

# Permutation p-value
p_perm <- mean(c(perm_F, obs_F) >= obs_F)
cat(
  "Observed F:", round(obs_F, 3),
  "\nPermutation p-value:", round(p_perm, 3), "\n"
  )
```

```{r}
#| fig-cap: "Permutation distribution of the $F$ statistic under the null hypothesis (4999 permutations). The red dashed line marks the observed $F$ statistic. The permutation p-value is the proportion of permuted F values to the right of this line."
#| cache: true
#| label: fig-permutation-plot

perm_df <- data.frame(F = c(perm_F, obs_F))

ggplot(perm_df, aes(x = F)) +
  geom_histogram(binwidth = 0.2, fill = "#2E86AB",
                 colour = "white", alpha = 0.8) +
  geom_vline(xintercept = obs_F,
             colour = "red", linetype = "dashed",
             linewidth = 1) +
  annotate("text", x = obs_F + 0.3, y = Inf,
           label = paste0("Observed F = ", round(obs_F, 2)),
           colour = "red", hjust = 0, vjust = 2, size = 3.5) +
  labs(x = "F statistic", y = "Count") +
  theme_bw()
```

Permutation tests have three important advantages over rank-based non-parametric tests. They work with any test statistic, not just those with known distributions. They make no assumptions about the shape of the residual distribution. And they are exact in the sense that their type I error rate is exactly $\alpha$ in finite samples, whereas rank tests rely on large-sample approximations.

Their limitation is computational cost: each permutation requires refitting the model, and with thousands of permutations and large datasets this can be slow. Parallelisation with `furrr`, as shown above, mitigates this considerably.

## Robust ANOVA with WRS2

Robust statistical methods are a middle ground between classical ANOVA and non-parametric tests. Rather than discarding the distributional framework entirely, they modify it to be less sensitive to outliers and heavy tails by replacing the sample mean with more resistant location estimators like trimmed means or M-estimators.

The `WRS2` package [@Mair2020] implements a comprehensive suite of robust ANOVA methods in R.

### One-Way Robust ANOVA

```{r robust-anova}
#| cache: true
# install.packages("WRS2") if needed
library(WRS2)

set.seed(42)

# Data with outliers — classical ANOVA will be distorted
y_out  <- c(rnorm(15, mean = 20, sd = 3),
            rnorm(15, mean = 25, sd = 3),
            rnorm(15, mean = 22, sd = 3))

# Inject two extreme outliers into the control group
y_out[c(3, 7)] <- c(55, 62)

group_out <- factor(rep(c("Control", "Nitrogen", "Phosphorus"),
                        each = 15))
df_out    <- data.frame(y = y_out, group = group_out)

# Classical ANOVA — affected by outliers
summary(aov(y ~ group, data = df_out))

# Robust one-way ANOVA using trimmed means (tr = 0.2 = 20% trimming)
t1way(y ~ group, data = df_out, tr = 0.2)

# Post-hoc comparisons with robust method
lincon(y ~ group, data = df_out, tr = 0.2)
```

The `t1way()` function tests for differences in trimmed means across groups, using a heteroscedastic test statistic that is robust to both outliers and unequal variances simultaneously. The trimming proportion `tr = 0.2` removes the most extreme 20% of observations from each tail before computing the mean. This standard choice provides substantial robustness while retaining reasonable efficiency.

### Robust Repeated Measures ANOVA

```{r robust-rm}
#| cache: true
set.seed(42)

n_subj <- 20

# Repeated measures data with outliers in long format
df_rm_out <- data.frame(
  y       = c(rnorm(n_subj, 80, 8),
              rnorm(n_subj, 75, 8),
              rnorm(n_subj, 70, 8),
              rnorm(n_subj, 68, 8)),
  subject = factor(rep(1:n_subj, times = 4)),
  time    = factor(rep(1:4, each = n_subj))
)

# Inject two extreme outliers
df_rm_out$y[c(5, 38)] <- c(130, 20)

# Robust repeated measures ANOVA
# rmanova() requires explicit vectors, not a formula
WRS2::rmanova(
  y      = df_rm_out$y,
  groups = df_rm_out$time,
  blocks = df_rm_out$subject,
  tr     = 0.2
)

# Post-hoc robust pairwise comparisons
WRS2::rmmcp(
  y      = df_rm_out$y,
  groups = df_rm_out$time,
  blocks = df_rm_out$subject,
  tr     = 0.2
)
```

### When to Use Robust Methods

Robust ANOVA methods are particularly valuable when:

- Outliers are genuine biological observations. Removing them is not scientifically justified, but their influence on the mean should be limited.
- The response distribution is heavy-tailed but not severely skewed. In such case, the trimmed mean is still a meaningful location estimator.
- You want results that are directly interpretable in terms of location differences, unlike rank-based tests.

Their main limitation is that trimmed mean differences are less familiar to most biological audiences than ordinary mean differences, and reporting them requires more explanation.

## Generalised Linear Models as the Real Solution

### The Fundamental Problem with Transformations

All of the methods discussed so far, transformations, rank tests, robust ANOVA, are __workarounds__ applied to a model that was not designed for the data at hand. When the response is a count, a proportion, a binary outcome, or a survival time, the normal linear model is structurally misspecified: it assumes a symmetric, continuous, unbounded response, while the actual response is discrete, bounded, or strictly positive. No transformation or rank-based adjustment changes this fundamental mismatch.

The correct solution is to choose a model whose structure matches the data-generating process. **Generalised linear models (GLMs)** provide exactly this: a unified framework that extends the linear model to non-normal response distributions through two modifications.

### The GLM Framework {#sec-glm-intro}

A GLM has three components:

1. A **random component**: the distribution of the response, chosen from the exponential family (normal, Poisson, binomial, gamma, negative binomial, etc.).
2. A **systematic component**: the linear predictor $\eta_i = \mathbf{x}_i^T \boldsymbol{\beta}$, the same linear combination of predictors as in the normal linear model.

::: {.callout-tip}
## Show me the maths!
Found $\eta_i = \mathbf{x}_i^T \boldsymbol{\beta}$, the bold symbols and superscript $T$ unfamiliar, We have you covered! See @sec-matrix-notation for a simple and complete explanation.
:::

3. A **link function** $g(\cdot)$ connecting the expected response to the linear predictor: $g(\mu_i) = \eta_i$.

The model is fitted by maximum likelihood rather than ordinary least squares, and inference is based on likelihood ratio tests and Wald tests rather than F tests.

### Common GLMs in Biology

| Response type | Distribution | Link function | R family |
|---------------|-------------|---------------|----------|
| Counts (no overdispersion) | Poisson | log | `poisson` |
| Counts (overdispersed) | Negative binomial | log | `MASS::negative.binomial` |
| Proportions | Binomial | logit | `binomial` |
| Binary outcomes | Binomial | logit | `binomial` |
| Positive continuous | Gamma | log or inverse | `Gamma` |
| Positive continuous (log-normal) | Gaussian | log | `gaussian(link="log")` |

### Example: Count Data

Suppose you are counting the number of insect damage events on plants under three pesticide treatments. The response is a count, non-negative integers with variance likely proportional to the mean. The correct model is a Poisson GLM or, if the counts are overdispersed, a negative binomial GLM.

```{r glm-count}
#| cache: true
#| message: false
library(MASS)

set.seed(42)

n <- 30
group <- factor(rep(c("Control", "Low dose", "High dose"), each = n),
                  levels = c("Control", "Low dose", "High dose"))
counts <- c(rpois(n, lambda = 15),
             rpois(n, lambda = 8),
             rpois(n, lambda = 3))
df_glm <- data.frame(counts, group)

# Wrong approach: ANOVA on raw counts
summary(aov(counts ~ group, data = df_glm))

# Correct approach: Poisson GLM
fit_pois <- glm(counts ~ group, data = df_glm,
                family = poisson(link = "log"))
summary(fit_pois)

# Check for overdispersion
cat("Residual deviance / df:",
    round(deviance(fit_pois) / df.residual(fit_pois), 3), "\n")

# If overdispersed (ratio >> 1), use negative binomial
fit_nb <- glm.nb(counts ~ group, data = df_glm)
summary(fit_nb)

# ANOVA-style table for GLM
car::Anova(fit_pois, type = "III")
car::Anova(fit_nb,   type = "III")
```

Both the ANOVA and the Poisson GLM detect a highly significant group effect, but comparing them reveals important differences in what each analysis is actually doing and how trustworthy its results are.

The standard ANOVA ($F_{2,87} = 150.9$, $p < 0.001$) treats the counts as if they were continuous normally distributed measurements. It finds a significant result, but its residual mean square (8.8) has no meaningful interpretation for count data, and the $F$ distribution used to compute the p-value is only an approximation whose quality depends on assumptions that are structurally wrong for integer-valued responses.

The Poisson GLM produces coefficients on the log scale that have a direct multiplicative interpretation. The intercept ($\hat{\beta}_0 = 2.771$) is the log of the expected count in the Control group: $e^{2.771} \approx 16$ which is close to the true $\lambda = 15$. The Low dose coefficient ($\hat{\beta} = -0.849$) gives a rate ratio of $e^{-0.849} \approx 0.43$: the Low dose group has approximately 43% of the count of the Control group. The High dose coefficient gives $e^{-1.661} \approx 0.19$: the High dose group has only about 19% of the Control count. These are the scientifically natural quantities for count data, rate ratios, not arithmetic differences.

The overdispersion ratio of 1.147 is reassuringly close to 1, indicating that the Poisson model is adequate for these data and overdispersion is not a concern here. This is expected since the data were generated from a Poisson distribution. The negative binomial model confirms this: its estimated dispersion parameter $\hat{\theta} = 37{,}871$ is enormous, meaning the negative binomial has converged to a Poisson (the negative binomial approaches Poisson as $\theta \to \infty$). The iteration limit warning and the unstable standard error on $\hat{\theta}$ ($SE = 1{,}198{,}071$) are further confirmation that the extra dispersion parameter is not needed indicating that the Poisson model is sufficient. The AIC for the Poisson model (435.75) is lower than for the negative binomial (437.75), correctly penalising the unnecessary extra parameter.

The ANOVA-style deviance table from `car::Anova()` gives a likelihood ratio test for the group effect: $\chi^2_2 = 306.81$, $p < 0.001$ for the Poisson model, essentially identical to the negative binomial result ($\chi^2_2 = 306.74$). This is the appropriate test statistic for a GLM, not an $F$ statistic, but a likelihood ratio chi-squared, and it is this value, not the ANOVA $F$ statistic, that should be reported when analysing count data with a Poisson GLM.

### Example: Proportion Data

```{r glm-proportion}
set.seed(42)

# Number of plants showing disease symptoms out of n_total
n_total   <- 20
n_disease <- c(rbinom(10, n_total, prob = 0.7),   # Control
               rbinom(10, n_total, prob = 0.4),   # Low dose
               rbinom(10, n_total, prob = 0.15))  # High dose
group_p <- factor(rep(c("Control", "Low dose", "High dose"), each = 10),
                    levels = c("Control", "Low dose", "High dose"))

df_prop <- data.frame(
  n_disease = n_disease,
  n_healthy = n_total - n_disease,
  group = group_p
)

# Correct approach: binomial GLM
# cbind(successes, failures) for proportion data
fit_binom <- glm(cbind(n_disease, n_healthy) ~ group,
                 data   = df_prop,
                 family = binomial(link = "logit"))
summary(fit_binom)
car::Anova(fit_binom, type = "III")

# Back-transform coefficients to probability scale
# using the inverse logit
pred_df <- data.frame(group = levels(group_p))
pred_df$prob <- predict(fit_binom,
                        newdata = pred_df,
                        type = "response")
print(pred_df)
```

The binomial GLM produces a clear and directly interpretable result.

The intercept ($\hat{\beta}_0 = 0.641$) is the log-odds of disease in the Control group. Back-transforming via the inverse logit gives $\text{plogis}(0.641) = 0.655$, a 65.5% probability of showing disease symptoms in the control group, consistent with the true simulation probability of 0.70. The Low dose coefficient ($\hat{\beta} = -0.903$) is a log-odds ratio: exponentiated, $e^{-0.903} = 0.41$, meaning the odds of disease in the Low dose group are 59% lower than in the Control group. The High dose coefficient gives $e^{-2.091} = 0.12$, an 88% reduction in the odds of disease relative to Control. These are the scientifically natural quantities for proportion data: odds ratios and predicted probabilities, not arithmetic differences in means.

The back-transformed predicted probabilities make the dose-response relationship immediately readable: disease probability declines from 65.5% in the Control group to 43.5% under low dose treatment and to 19.0% under high dose treatment, a near-linear decline in probability across the three groups that the model has estimated honestly on the logit scale without any transformation of the response.

The residual deviance (27.49 on 27 degrees of freedom) is encouragingly close to its expected value under a well-fitting Poisson or binomial model and the ratio of 1.02 gives no indication of overdispersion. Had this ratio been substantially greater than 1, a quasi-binomial model or a beta-binomial model (Section 11.3) would be warranted.

The likelihood ratio test from `car::Anova()` gives $\chi^2_2 = 92.75$, $p < 0.001$, confirming that pesticide treatment has a highly significant effect on disease incidence. As with the count example, this chi-squared statistic, not an $F$ statistic, is the appropriate test for a binomial GLM and is what should be reported in the methods section of a paper.

A complete result report reads:

> The effect of pesticide dose on disease incidence was analysed
> using a binomial GLM with a logit link. Pesticide treatment had
> a significant effect on the probability of disease symptoms
> ($\chi^2_2 = 92.75$, $p < 0.001$). Predicted disease probabilities
> were 65.5% (Control), 43.5% (Low dose), and 19.0% (High dose).
> Relative to the Control group, Low dose reduced the odds of
> disease by 59% (OR = 0.41, 95% CI: ...) and High dose reduced
> them by 88% (OR = 0.12, 95% CI: ...).

### GLMs vs Transformations: A Direct Comparison

```{r glm-vs-transform}
#| cache: true
#| fig-cap: "Comparison of three approaches to analysing count data: ANOVA on raw counts (top left), ANOVA on log-transformed counts (top right), and a Poisson GLM (bottom). The Poisson GLM produces the most appropriate residual structure — the Q-Q plot is closest to the diagonal — because it models the data-generating process directly rather than approximating it through a transformation."
#| label: fig-glm-vs-transform
fit_raw_aov <- aov(counts ~ group, data = df_glm)
fit_log_aov <- aov(log(counts + 1) ~ group, data = df_glm)

res_df <- data.frame(
  residuals = c(residuals(fit_raw_aov),
                residuals(fit_log_aov),
                residuals(fit_pois, type = "pearson")),
  model = rep(c("ANOVA (raw)", 
                "ANOVA (log)",
                "Poisson GLM"),
                each = nrow(df_glm))
)

ggplot(res_df, aes(sample = residuals)) +
  stat_qq(colour = "#2E86AB", alpha = 0.7) +
  stat_qq_line(colour = "#E84855", linewidth = 0.8) +
  facet_wrap(~ model) +
  labs(x = "Theoretical quantiles", y = "Sample quantiles") +
  theme_bw()
```

### A Decision Framework

```{mermaid}
%%| label: fig-assumption-failure-tree
%%| fig-cap: "Decision framework for choosing an analysis strategy when ANOVA assumptions are violated."

flowchart TD
    A[Assumption\nviolation detected] --> B{What type\nof response?}

    B -- Continuous\npositive --> C{Multiplicative\nvariation?}
    B -- Count data --> D[Poisson or\nNegative Binomial GLM]
    B -- Proportion\nor binary --> E[Binomial GLM]
    B -- Continuous\nunbounded --> F{Outliers or\nheavy tails?}

    C -- Yes --> G[Log transform\nor Gamma GLM]
    C -- No --> H[Square root\nor robust ANOVA]

    F -- Outliers --> I[Robust ANOVA\nWRS2]
    F -- Heavy tails --> J[Permutation test\nor Kruskal-Wallis]
    F -- Neither --> K[Proceed with\nANOVA]

    style A fill:#555555,color:#ffffff
    style B fill:#F6AE2D,color:#ffffff
    style C fill:#2E86AB,color:#ffffff
    style D fill:#81B29A,color:#ffffff
    style E fill:#81B29A,color:#ffffff
    style F fill:#2E86AB,color:#ffffff
    style G fill:#81B29A,color:#ffffff
    style H fill:#81B29A,color:#ffffff
    style I fill:#81B29A,color:#ffffff
    style J fill:#81B29A,color:#ffffff
    style K fill:#81B29A,color:#ffffff
```

### The Bottom Line

The hierarchy of approaches for dealing with assumption violations reflects an important principle: **the further a remedy is from the data-generating process, the less satisfying it is**. A log transformation is a reasonable approximation for multiplicative data; a Poisson GLM is the correct model. A rank test avoids distributional assumptions; a GLM with the right family makes the right assumptions. Robust ANOVA limits the influence of extreme values; a heavy-tailed distribution models them explicitly.

For continuous responses with approximately normal residuals and minor violations, ANOVA and its extensions remain the right tools. For discrete, bounded, or structurally non-normal responses, the investment in learning GLMs pays dividends not only in statistical correctness but in interpretability: a log-odds ratio from a logistic regression, a rate ratio from a Poisson regression, or a mean ratio from a gamma regression all have direct biological meaning that a rank difference or a trimmed mean difference does not.